---
title: "caseStudy2"
author: "Tadd Backus"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction and Overview
In this report, we explore the employee attrition dataset. This dataset contains
employees who have either left a company or not as well as 37 different variables
potentially relating to this. The steps and procedures taken for this analysis 
are detailed below. We start by looking at the most important factors for employee
attrition and then any trends for different job roles. Finally, we create two 
prediction models, one classifier to predict attrition and one linear regression
to predict monthly income.

# Useful Links

* [`GitHub`](https://github.com/taddbackus/CaseStudy2DDS):
This link contains all of the data, R code, and power points.

* ['Employee Attrition App'](https://tbackus.shinyapps.io/EmployeeAttrition/?_ga=2.59679213.1944325347.1659738568-2039603706.1659738568):
This link contains an R Shiny app that allows the user to explore the attrition rates 
of different job roles and education fields, as well as the monthly income.

* ['Presentation of Results - YouTube'](https://www.youtube.com/watch?v=iC9egHIgbzw):
Video of the presentation of results of Employee Attrition analysis.

* [`README.md`](https://github.com/tadbackus/CaseStudy2DDS/blob/main/README.md): 
This page shows an introduction to the analysis we ran. It also includes a few 
details about the datasets provided.

* [`Beers.csv`](https://github.com/tadbackus/CaseStudy2DDS/blob/main/CaseStudy2-data.csv) 
 



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load packages
library(aws.s3)
library(RCurl)
library(class)
library(caret)
library(e1071)
library(tidyr)
library(tidyverse)
library(olsrr)
library(forecast)

```

```{r, echo=FALSE}
attrData = read.csv('/Users/taddbackus/School/summer22/doingDataScience/caseStudy2/CaseStudy2-data.csv',header=TRUE)
attrTest = read.csv('/Users/taddbackus/School/summer22/doingDataScience/caseStudy2/CaseStudy2CompSet No Attrition.csv',header=TRUE)
miTest = read.csv('/Users/taddbackus/School/summer22/doingDataScience/caseStudy2/CaseStudy2CompSet No Salary.csv',header=TRUE)

options(scipen=999)  

aDF = attrData
aDF$Department = factor(aDF$Department)
aDF$EducationField = factor(aDF$EducationField)
aDF$BusinessTravel = factor(aDF$BusinessTravel)
aDF$Gender = factor(aDF$Gender)
aDF$JobRole = factor(aDF$JobRole)
aDF$MaritalStatus = factor(aDF$MaritalStatus)
aDF$OverTime = factor(aDF$OverTime)
aDF$Attrition = factor(aDF$Attrition)

attrY <- aDF %>% filter(Attrition=='Yes')
attrN <- aDF %>% filter(Attrition=='No')
```

### This section looks at factors that have a large impact on employee attrition.  
The bar graph below shows the amount of employees that left and stayed for each of the departments.
```{r, echo=FALSE}
# Attrition Analysis
# Job Role/Department
ggplot(aDF,aes(x=Department,fill=Attrition))+
  geom_bar()+
  coord_flip()+
  ggtitle('Attrition by Department')+
  ylab('Count')+
  xlab('Department')
```

Below are the numbers for employees who left.
```{r, echo=FALSE}
summary(attrY$Department)
```
Below are the numbers for employees who stayed
```{r, echo=FALSE}
summary(attrN$Department)
```

The bar graph below shows the amount of employees that left and stayed for each individual job role.
```{r, echo=FALSE}
ggplot(aDF,aes(x=JobRole,fill=Attrition))+
  geom_bar()+
  coord_flip()+
  ggtitle('Attrition by Job Role')+
  ylab('Count')+
  xlab('Job Role')
```

Below are the numbers for employees who left.
```{r, echo=FALSE}
summary(attrY$JobRole)
```
Below are the numbers for employees who stayed.
```{r, echo=FALSE}
summary(attrN$JobRole)
```

This plot shows combinations of education field and job role and whether or not the employee left.
```{r, echo=FALSE}
ggplot(aDF,aes(x=EducationField,y=JobRole,color=Attrition))+
  geom_point(position='jitter')+
  ggtitle('Attrition of Job Roles by Education Field')+
  xlab('Education Field')+
  ylab('Job')
```

These graphs show the distributions of log transformed monthly incomes for the corresponding attrition.
```{r, echo=FALSE}
# Monthly Income
ggplot(aDF,aes(x=log(MonthlyIncome),y=Attrition,fill=Attrition))+
  geom_boxplot()+
  ggtitle('Attrition based on Monthly Income')+
  xlab('Monthly Income')

ggplot(aDF,aes(x=log(MonthlyIncome),fill=Attrition))+
  geom_histogram(bins=20)+
  facet_wrap(~Attrition)+
  ggtitle('Histogram of Monthly Incomes')+
  xlab('Monthly Income')+
  ylab('Count')
```

The results of a two sample t-test on the monthly income data.
```{r, echo=FALSE}
t.test(x=log(attrY$MonthlyIncome), y=log(attrN$MonthlyIncome), conf.int=.95, var.equal=TRUE, alternative="two.sided")
```
There is evidence to suggest that the median income for workers who quit is lower than workers who did not quit
(p-value<0.0000001)

This barplot shows how many employees left based on whether or not they worked over time.
```{r, echo=FALSE}
# Over Time
ggplot(aDF,aes(x=OverTime,fill=Attrition))+
  geom_bar()+
  ggtitle('Attrition for Over Time workers')+
  xlab('Count')+
  ylab('Over Time status')
```

Below are the numbers for employees who left.
```{r, echo=FALSE}
summary(attrY$OverTime)
```
Below are the numbers for employees who stayed.
```{r, echo=FALSE}
summary(attrN$OverTime)
```

This boxplot shows the distribution of monthly incomes for each job role.
```{r, echo=FALSE}
# Job Role Analysis
# Monthly Income
ggplot(aDF,aes(x=MonthlyIncome,y=JobRole,fill=Attrition))+
  geom_boxplot()+
  ggtitle('Monthly Income for Job Roles')+
  xlab('Monthly Income')+
  ylab('Job Role')
```

Plotting job satisfaction for each job role.
```{r, echo=FALSE}
# Job Satisfaction
ggplot(aDF,aes(x=JobSatisfaction,y=JobRole,color=Attrition))+
  geom_point(position='jitter')+
  ggtitle('Job Role and Satisfaction')+
  xlab('Job Satisfaction')+
  ylab('Job Role')
```


```{r, echo=FALSE}
attrModel = attrTest
happyKnn = subset(aDF, select=c(Attrition,Age,DistanceFromHome,EnvironmentSatisfaction,
                                JobSatisfaction,RelationshipSatisfaction,WorkLifeBalance))
payKnn = subset(aDF, select=c(Attrition,Age,DailyRate,HourlyRate,MonthlyIncome,MonthlyRate,
                              PercentSalaryHike,PerformanceRating,StockOptionLevel))
yearKnn = subset(aDF, select=c(Attrition,Age,Education,JobLevel,JobInvolvement,JobSatisfaction,
                               NumCompaniesWorked,TotalWorkingYears,TrainingTimesLastYear,
                               YearsAtCompany,YearsInCurrentRole,YearsSinceLastPromotion,
                               YearsWithCurrManager))
happyNB = subset(aDF, select=c(Attrition,Age,DistanceFromHome,
                               EnvironmentSatisfaction,JobSatisfaction,
                               RelationshipSatisfaction,WorkLifeBalance,
                               Gender,JobRole,OverTime,MaritalStatus))
payNB = subset(aDF, select=c(Attrition,Age,DailyRate,HourlyRate,MonthlyIncome,
                             MonthlyRate,PercentSalaryHike,PerformanceRating,
                             StockOptionLevel,Department,EducationField,
                             Gender,JobRole))
yearNB = subset(aDF, select=c(Attrition,Age,Education,JobLevel,JobInvolvement,
                              JobSatisfaction,NumCompaniesWorked,TotalWorkingYears,
                              TrainingTimesLastYear,YearsAtCompany,YearsInCurrentRole,
                              YearsSinceLastPromotion,YearsWithCurrManager))
```

Variables used in the "Happiness" kNN Model.
```{r, echo=FALSE}
summary(happyKnn)
```

Variables used in the "Pay" kNN Model.
```{r, echo=FALSE}
summary(payKnn)
```

Variables used in the "Experience" kNN Model.
```{r, echo=FALSE}
summary(yearKnn)
```

Variables used in the "Happiness" Naive Bayes Model.
```{r, echo=FALSE}
summary(happyNB)
```

Variables used in the "Pay" Naive Bayes Model.
```{r, echo=FALSE}
summary(payNB)
```

Variables used in the "Experience" Naive Bayes Model.
```{r, echo=FALSE}
summary(yearNB)
```

```{r, echo=FALSE}
happyKnnTest = subset(attrModel, select=c(Age,DistanceFromHome,EnvironmentSatisfaction,
                                JobSatisfaction,RelationshipSatisfaction,WorkLifeBalance))
payKnnTest = subset(attrModel, select=c(Age,DailyRate,HourlyRate,MonthlyIncome,MonthlyRate,
                              PercentSalaryHike,PerformanceRating,StockOptionLevel))
yearKnnTest = subset(attrModel, select=c(Age,Education,JobLevel,JobInvolvement,JobSatisfaction,
                               NumCompaniesWorked,TotalWorkingYears,TrainingTimesLastYear,
                               YearsAtCompany,YearsInCurrentRole,YearsSinceLastPromotion,
                               YearsWithCurrManager))
happyNBTest = subset(attrModel, select=c(Age,DistanceFromHome,
                               EnvironmentSatisfaction,JobSatisfaction,
                               RelationshipSatisfaction,WorkLifeBalance,
                               Gender,JobRole,OverTime,MaritalStatus))
payNBTest = subset(attrModel, select=c(Age,DailyRate,HourlyRate,MonthlyIncome,
                             MonthlyRate,PercentSalaryHike,PerformanceRating,
                             StockOptionLevel,Department,EducationField,
                             Gender,JobRole))
yearNBTest = subset(attrModel, select=c(Age,Education,JobLevel,JobInvolvement,
                              JobSatisfaction,NumCompaniesWorked,TotalWorkingYears,
                              TrainingTimesLastYear,YearsAtCompany,YearsInCurrentRole,
                              YearsSinceLastPromotion,YearsWithCurrManager))

happyKnnMod = train(happyKnn[,2:7],happyKnn$Attrition,method='knn')
payKnnMod = train(payKnn[,2:9],payKnn$Attrition,method='knn')
yearKnnMod = train(yearKnn[,2:13],yearKnn$Attrition,method='knn')

happyNBMod = naiveBayes(happyNB[,2:11],happyNB$Attrition)
payNBMod = naiveBayes(payNB[,2:13],payNB$Attrition)
yearNBMod = naiveBayes(yearNB[,2:13],yearNB$Attrition)

attrModel$happyKnnPred = predict(object=happyKnnMod,happyKnnTest[,1:6])
attrModel$payKnnPred = predict(object=payKnnMod,payKnnTest[,1:8])
attrModel$yearKnnPred = predict(object=yearKnnMod,yearKnnTest[,1:12])
attrModel$happyKnnProb = predict(object=happyKnnMod,happyKnnTest[,1:6],type='prob')
attrModel$payKnnProb = predict(object=payKnnMod,payKnnTest[,1:8],type='prob')
attrModel$yearKnnProb = predict(object=yearKnnMod,yearKnnTest[,1:12],type='prob')

attrModel$happyNBPred = predict(happyNBMod,happyNBTest[,1:10])
attrModel$payNBPred = predict(payNBMod,payNBTest[,1:12])
attrModel$yearNBPred = predict(yearNBMod,yearNBTest[,1:12])
attrModel$happyNBProb = predict(happyNBMod,happyNBTest[,1:10],'raw')
attrModel$payNBProb = predict(payNBMod,payNBTest[,1:12],'raw')
attrModel$yearNBProb = predict(yearNBMod,yearNBTest[,1:12],'raw')

attrModel$probAvgNB = (attrModel$happyNBProb[,'Yes']+attrModel$payNBProb[,'Yes']+attrModel$yearNBProb[,'Yes'])/3
attrModel$probWeightNB = attrModel$happyNBProb[,'Yes']*.4+attrModel$payNBProb[,'Yes']*.2+attrModel$yearNBProb[,'Yes']*.4

attrModel$probAvgKnn = (attrModel$happyKnnProb$Yes+attrModel$payKnnProb$Yes+attrModel$yearKnnProb$Yes)/3
attrModel$probWeightKnn = attrModel$happyKnnProb$Yes*.6+attrModel$payKnnProb$Yes*.2+attrModel$yearKnnProb$Yes*.2

attrModel$probAvg = (attrModel$probAvgNB+attrModel$probAvgKnn)/2
attrModel$probWeight = (attrModel$probWeightNB+attrModel$probWeightKnn)/2

attrModel$predAvg = as.factor(ifelse(attrModel$probAvg>0.5,'Yes','No'))
attrModel$predWeight = as.factor(ifelse(attrModel$probWeight>0.5,'Yes','No'))

classPred = subset(attrModel, select=c(ID,predWeight))
```

Combined model = (weighted kNN + weighted NB)/2
Weighted kNN = happiness X 0.6 + pay X 0.2 + experience X 0.2
Weight NB = happines X 0.4 + pay X 0.2 + experience X 0.4

Creating the linear regression model to predict monthly income.
Variables used in the model.
```{r, echo=FALSE}
miModel = miTest
fullFit = lm(MonthlyIncome~JobLevel+JobRole+TotalWorkingYears+DistanceFromHome+
               Age+PerformanceRating+YearsAtCompany+YearsSinceLastPromotion+OverTime,data=aDF)
summary(fullFit)
MonthlyIncome = predict(fullFit,newdata=miModel)
miPred = cbind(miModel,MonthlyIncome)
miPred = subset(miPred,select=c(ID,MonthlyIncome))
```
Performance of the model
```{r, echo=FALSE}
CV(fullFit)
```
RMSE of the linear regression model
```{r,echo = FALSE}
sqrt(mean(fullFit$residuals^2))
```
Plot of fitted values for monthly income and the originals
```{r, echo=FALSE}
aDF$lmPred = fullFit$fitted.values
ggplot(aDF,aes(x=lmPred,y=MonthlyIncome))+
  geom_point()+
  geom_smooth(method='lm')+
  ggtitle('Linear Regression for Monthly Income')+
  xlab('Predicted Values')+
  ylab('Monthly Income')

